---
title: hw5
date: "4/15/26"
format: pdf
---
[Download source](https://raw.githubusercontent.com/ywanglab/STAT3000/refs/heads/main/psets/hw5.qmd)

This problem set explores probability through two settings: repeated dice rolls and American roulette. You will use both Monte Carlo simulation and exact mathematical calculations to understand these models.

Please answer each of the exercises below. For those asking for a mathematical calculation, use LaTeX to show your work.

Important: Make sure that your document renders in less than 5 minutes.

## Part I: Rolling two dice

1. Write a function called `double_six` that takes a number `n` as an argument, simulates rolling two fair dice `n` times, and returns `TRUE` if at least one of the `n` rolls is a double-six.

   Hint: generate two vectors of rolls and check whether any pair equals `(6, 6)`.

```{r}
double_six <- function(n){
  ## Your code here
  die1 <- sample(1:6,n,replace = TRUE)
  die2 <- sample(1:6,n,replace = TRUE)
  any(die1 == 6 & die2 == 6)}
```

2. Suppose you roll two fair dice 24 times. If we define success to mean seeing at least one double-six, what is the estimated probability of success? Use a Monte Carlo simulation with `B = 1000` trials based on the function `double_six` from the previous exercise.

```{r}
B <- 10^3
n <- 24
## Your code here
results <- replicate(B,double_six(n))
mean(results)
```

3. Redo the previous exercise for several values of `n` to determine at what number of rolls the probability becomes greater than 50%. Set the seed at 1997.

```{r}
set.seed(1997)
compute_prob <- function(n, B = 10^3){
  mean(replicate(B,double_six(n)))}

n_vals <- 1:60
## Your code here
probs <- sapply(n_vals,compute_prob)
```

4. These probabilities can be computed exactly instead of relying on Monte Carlo approximations. Since the probability of not getting a double-six on a single roll is `35/36`, we have

$$
\Pr(\text{at least one double-six in } n \text{ rolls})
= 1 - \left(\frac{35}{36}\right)^n.
$$

Plot the probabilities you obtained using Monte Carlo as points and the exact probabilities with a red line.

```{r}
exact_prob <- function(n){
  ## Your code here
  1 - (35/36)^n}

## Your code here
exact_prob <- sapply(n_vals,exact_prob)
plot(n_vals,probs,
     pch = 16,
     xlab = "Number of Rolls (n)",
     ylab = "Probability",
     main = "Monte Carlo vs Exact Probability")
lines(n_vals,exact_prob,col = "red",lwd = 2)
```

5. The Monte Carlo points in question 4 will not exactly match the red curve because the simulation uses only 1,000 iterations. Repeat exercise 2 for `n = 24`, but now try `B <- seq(10, 250, 5)^2` iterations. Plot the estimated probability against `sqrt(B)`. At what value of `sqrt(B)` do the estimates consistently stay within `0.005` of the exact probability? Add horizontal lines at the exact probability plus and minus `0.005`. Set the seed to 1998.

```{r}
set.seed(1998)
B <- seq(10, 250, 5)^2
n <- 24
## Your code here
exact <- 1 - (35/36)^24
estimates <- sapply(B,function(b){
  mean(replicate(b,double_six(n)))})
```

6. Repeat the comparison from question 4 (Monte Carlo points versus exact red line), but now use your findings from question 5 to choose an appropriate value of `B` so that the points practically fall on the red curve.

   Hint: choose a value of `B` that is large enough for a good plot, but not so large that your document takes too long to render.

```{r}
n_vals <- 1:60
## Your code here
B <- 20000
probs <- sapply(n_vals, function(n){
  die1 <- matrix(sample(1:6, n * B, replace = TRUE), nrow = B)
  die2 <- matrix(sample(1:6, n * B, replace = TRUE), nrow = B)
  mean(rowSums(die1 == 6 & die2 == 6) > 0)})
```

## Part II: American roulette

In American roulette there are 38 slots total: 18 red, 18 black, and 2 green.

7. If a player bets $1 on red, what is the probability that the casino wins the bet?

$$
P(\text{casino wins}) = P(X = 1) = /frac{20}{38}
$$

8. If a player bets $1 on red, the casino's profit from that single bet is represented by the random variable $X$:

- $X = -1$ if the ball lands on red,
- $X = +1$ otherwise.

Create a sampling model for `X` using the `sample` function.

```{r}
## Your code here
X <- sample(c(-1,1),size = 1,replace = TRUE, prob = c(18/38,20/38))
replicate(10, sample(c(-1,1), size = 1, replace = TRUE, prob = c(18/38,20/38)))
```

9. Now create a random variable $S$ for the casino's total profit if `n = 1000` people each make a $1 bet on red. Use Monte Carlo simulation with `B = 10000` trials to estimate the probability that the casino loses money.

```{r}
n <- 1000
B <- 10^4
## Your code here
one_trial <- function(n){
  sum(sample(c(-1, 1), size = n, replace = TRUE,
             prob = c(18/38, 20/38)))}
S_vals <- replicate(B, 
  sum(sample(c(-1,1), size = n, replace = TRUE,
             prob = c(18/38, 20/38))))
mean(S_vals < 0)
```

10. What is the expected value of $X$?

$$
E[X] = (-1)\cdot \frac{18}{38} + (1)\cdot \frac{20}{38}
= \frac{-18 + 20}{38}
= \frac{2}{38}
= \frac{1}{19}
$$

11. What is the standard error of $X$?

$$
\mathrm{SE}(X) = \sqrt{\mathrm{Var}(X)} = \sqrt{E[X^2] - (E[X])^2}
$$

$$
E[X^2] = 1
$$

$$
\mathrm{Var}(X) = 1 - \left(\frac{1}{19}\right)^2
= 1 - \frac{1}{361}
= \frac{360}{361}
$$

$$
\mathrm{SE}(X) = \sqrt{\frac{360}{361}} = \frac{\sqrt{360}}{19}
$$

12. What is the expected value of $S$? Does the Monte Carlo simulation confirm this?

$$
S = X_1 + X_2 + \cdots + X_{1000}
$$

$$
E[S] = \sum_{i=1}^{1000} E[X_i] = 1000 \cdot E[X]
$$

$$
E[S] = 1000 \cdot \frac{1}{19} = \frac{1000}{19} \approx 52.63
$$

```{r}
## Your code here
set.seed(123)
n <- 1000
B <- 10^4
S_vals <- replicate(B, {
  sum(sample(c(-1, 1), size = n, replace = TRUE,
             prob = c(18/38, 20/38)))})
mean(S_vals)
```

13. What is the standard error of $S$? Does the Monte Carlo simulation confirm this?

$$
S = X_1 + X_2 + \cdots + X_{1000}
$$

$$
\mathrm{Var}(S) = \sum_{i=1}^{1000} \mathrm{Var}(X_i)
= 1000 \cdot \mathrm{Var}(X)
$$

$$
\mathrm{Var}(X) = \frac{360}{361}
$$

$$
\mathrm{SE}(S) = \sqrt{1000 \cdot \frac{360}{361}}
$$

```{r}
## Your code here
set.seed(123)
n <- 1000
B <- 10^4
S_vals <- replicate(B, {
  sum(sample(c(-1, 1), size = n, replace = TRUE,
             prob = c(18/38, 20/38)))})
sd(S_vals)
```

14. Use data visualization to convince yourself that the distribution of $S$ is approximately normal. Make a histogram and a QQ-plot of the standardized values of $S$. The QQ-plot should be close to the identity line.

```{r}
## Your code here
set.seed(123)
n <- 1000
B <- 10^4
S_vals <- replicate(B, {
  sum(sample(c(-1, 1), size = n, replace = TRUE,
             prob = c(18/38, 20/38)))})
mu_S <- n * (1/19)
se_S <- sqrt(n * (360/361))
Z <- (S_vals - mu_S) / se_S
```

15. The normal approximation will not be perfect in the tails. What would improve the approximation more: increasing the number of people playing `n`, or increasing the number of Monte Carlo iterations `B`? Explain your reasoning.

**Answer here**
```
n improves how well the distribution of 
S is approximated by a normal distribution.
B only improves how accurately you estimate or visualize that distribution.
```

16. Use the Central Limit Theorem to approximate the probability that the casino loses money. Does your CLT approximation agree with the Monte Carlo estimate from question 9?

$$
P(S < 0) = P\left(\frac{S - E[S]}{\mathrm{SE}(S)} < \frac{-E[S]}{\mathrm{SE}(S)}\right)
$$

$$
P(S < 0) \approx \Phi\left(\frac{-E[S]}{\mathrm{SE}(S)}\right)
$$

```{r}
## Your code here
n <- 1000
mu_X <- 1/19
var_X <- 360/361
mu_S <- n * mu_X
se_S <- sqrt(n * var_X)
clt_prob <- pnorm(0, mean = mu_S, sd = se_S)
clt_prob
```

17. What is the minimum number of people `n` who must each bet $1 on red so that the probability the casino loses money is approximately 1% according to the CLT? Check your answer with a Monte Carlo simulation.

$$
\text{Your derivation here}
$$

```{r}
## Your code here
mu_X <- 1/19
sigma_X <- sqrt(360/361)
z <- qnorm(0.01)
n_min <- (z * sigma_X / mu_X)^2
n_min
set.seed(123)
B <- 10000
S_vals <- replicate(B, {
  sum(sample(c(-1, 1), size = n_min, replace = TRUE,
             prob = c(18/38, 20/38)))})
mean(S_vals < 0)
```